---
title: "Super simple power analysis for 2024 data"
author: "Emma Cartuyvels"
date: "2025-05-05"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---
# Intro

In this document I attempt to make a simple simulation and power analysis using the simr package.

We use the following model:
$$
n = m + y + (1|l)
$$
With $n$ being the number of individuals (modeled using a Poisson distribution), $m$ the month, $y$ the year and $l$ the location. Since we are using a random effect for location we will be using the `simr` package which allows us to create an artificial mixed model object.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(here)
library(tidyverse)
library(simr)
```

# Information from SPRING experiment

At this moment we only have a limited data set to determine our coefficients from.

```{r gegevens-2024}
bestuivers_2024 <- read_csv(here("data", "processed", "bestuivers_2024.csv"))

bestuivers_2024_by_order <- bestuivers_2024 |>
  rename(location = location_code,
         month = maand) |>
  filter(
    species_nm != "Apis mellifera",
    method_cd == "TS"
  ) |>
  group_by(order, location, month) |>
  summarise(n_ind = sum(n_ind, na.rm = TRUE)) |>
  # correct for shorter transects compared to proposed protocol
  # length new transects / length old transects = (1000 * 3) / (160 * 4) ~= 5
  mutate(
    n_ind = n_ind * 5,
    month = as.factor(month)
  )
  

bestuivers_2024 <- bestuivers_2024 |>
  rename(location = location_code,
         month = maand) |>
  filter(
    species_nm != "Apis mellifera",
    method_cd == "TS"
  ) |>
  group_by(location, month) |>
  summarise(n_ind = sum(n_ind, na.rm = TRUE)) |>
  # correct for shorter transects compared to proposed protocol
  # length new transects / length old transects = (1000 * 3) / (160 * 4) ~= 5
  mutate(
    n_ind = n_ind * 5,
    month = as.factor(month)
  )
```

```{r glmer-model-pilotdata}
model1 <- glmer(
  n_ind ~ month + (1 | location),
  family = "poisson",
  data = bestuivers_2024
)
summary(model1)

sd_loc <- attr(VarCorr(model1)$location, "stddev")
sd_loc
```

```{r}
emmeans::emmeans(model1, ~ month)
```

```{r}
performance::check_overdispersion(model1)
performance::check_convergence(model1)
performance::check_outliers(model1)
performance::check_zeroinflation(model1)
performance::check_predictions(model1)
performance::check_group_variation(model1) |> summary()

```

```{r glmer-model-pilotdata-negbinom}
model2 <- glmer.nb(
  formula = n_ind ~ month + (1 | location),
  data = bestuivers_2024
)
summary(model2)

sd_loc <- attr(VarCorr(model2)$location, "stddev")
sd_loc
```

```{r}
emmeans::emmeans(model2, ~ month)
```

```{r}
performance::check_overdispersion(model2)
performance::check_convergence(model2)
performance::check_outliers(model2)
performance::check_predictions(model2)
performance::check_group_variation(model2) |> summary()
```




## Design Setup

```{r fun-generate-design}
# Define design variables
generate_design <- function(n_locations, n_years) {
  # Design variables
  years <- as.numeric(1:n_years)
  months <- as.factor(4:9)
  locs <- paste0("loc_", 1:n_locations)

  # Base: 1 visit per month per year per location
  design <- expand.grid(
    location = locs,
    year = years,
    month = months,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  design$month <- factor(design$month, levels = levels(months))
  return(design)
}
```

## Model Specification

```{r}
#' Fixed effects: Intercept (month 4), month 5-9, year (placeholder at index 7)
#' Month 4 and 5 based on SPRING data
fixed_base <- c(
  1.674776,
  0,
  unname(coef(summary(model1))[1, 1]) - 1.674776,
  unname(coef(summary(model1))[2:4, 1]) +
    unname(coef(summary(model1))[1, 1]) - 1.674776,
  NA
)

# Random effect variance for location and year
rand_loc <- attr(VarCorr(model1)$location, "stddev")
rand_year <- 0.15
```

## Power Analysis Function

```{r fun-run-power-analysis}
run_power_analysis <- function(
    year_slope_percent,
    n_locations,
    n_years,
    nsim = 100,
    fixed_effects = fixed_base,
    sd_rand_year,
    sd_rand_location = rand_loc
) {
  design <- generate_design(n_locations, n_years)
  
  # Convert percent trend to log scale
  year_slope <- log(1 + year_slope_percent / 100)
  fixed <- fixed_effects
  fixed[7] <- year_slope
  
  # random components
  var_corr <- list(
    location = c(sd_rand_location ^ 2),
    year = c(sd_rand_year ^ 2)
  )

  # Simulated model with defined structure
  model <- makeGlmer(
    formula = n_ind ~ month + year + (1 | location) + (1 | year),
    family = poisson,
    fixef = fixed,
    VarCorr = var_corr,
    data = design
  )

  test <- powerSim(
    model,
    nsim = nsim,
    # maybe change "z" (Wald test) to Likelihood ratio test ("lr")
    # lr is more precise, but will increase computation time
    # b/c model needs to be fitted twice, with/without year
    test = fixed("year", "lr"),
    progress = TRUE
  )

  return(test)
}
```

## Run Power Simulations

```{r run-simulations}
library(mirai)

# 1. Setup Daemons
n_cores <- parallel::detectCores(logical = FALSE)
daemons(n_cores, dispatcher = FALSE)

# 2. Define Parameter Grid (Flatten the loops)
trends <- c(1, 2)
locations <- c(15, 30, 60, 120, 240)
n_years_v  <- c(6, 12, 24)
sd_year <- c(0.01, 0.05, 0.15)

param_grid <- expand.grid(
  sdy = sd_year,
  ny = n_years_v,
  t = trends,
  l = locations,
  stringsAsFactors = FALSE
)

pr <- fs::dir_create(
  here::here("source", "powercalc_results")
)

# delete saved results?
delete <- FALSE
if (delete) fs::file_delete(pr)

# Store promises here
tasks <- list()

for (i in 1:nrow(param_grid)) {
  
  # Extract parameters for this specific row
  p <- param_grid[i, ]
  
  # Construct filename
  fname <- file.path(
    pr, 
    paste0("sdy", p$sdy ,"_ny", p$ny, "_trend", p$t, "_loc", p$l, ".rds")
  )
  
  # Submit to mirai
  tasks[[i]] <- mirai({
    library(here)
    library(lme4)
    library(simr)
    # Check cache
    if (file.exists(fname)) {
      res <- readRDS(fname)
      return(summary(res))
    }
    
    # Run Analysis
    result <- run_power_analysis(
      year_slope_percent = p$t,
      n_locations = p$l,
      n_years = p$ny,
      nsim = 100, # Per scenario
      fixed_effects = fixed_base,
      sd_rand_year = p$sdy,
      sd_rand_location = rand_loc
    )
    
    # Save and return powerSim object
    saveRDS(result, file = fname)
    return(summary(result)) 
  }, 
  # Pass parameters and functions
  environment()
  )
}

```



## Summarize Results

```{r collect-results}
# call_mirai blocks until all tasks are done
results_collected <- lapply(tasks, call_mirai)

# Attach results back to the grid
power_df <- param_grid |>
  bind_cols(
    lapply(results_collected, \(x) x$data) |> bind_rows()
  ) |>
  as_tibble() |>
  rename(
    sd_year = sdy,
    n_years = ny,
    trend = t,
    locations = l
  ) |>
  mutate(
    trend = factor(trend)
  )

# View Results
power_df

# Clean up
daemons(0)
```


## Extract and Plot Power Results

```{r plot-power-curves}
library(ggplot2)
ggplot(
  power_df,
  aes(x = locations, y = mean, color = trend)
) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(
    aes(
      ymin = lower,
      ymax = upper,
      fill = trend,
      colour = NULL
    ),
    alpha = 0.2
  ) +
  geom_point() +
  geom_hline(
    yintercept = 0.8,
    linetype = 2,
    color = "darkgrey"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(breaks = locations) +
  guides(x = guide_axis(check.overlap = TRUE)) +
  labs(
    title = "Power to Detect Yearly Trends",
    x = "Number of Locations",
    y = "Power"
  ) +
  facet_grid(
    sprintf("SD jaar: %.02f", sd_year) ~ sprintf("Aantal jaren: %02d", n_years)
  )
```

# Wild bees

```{r, eval=FALSE}
n_ind_per_loc <- bestuivers_2024 |>
  filter(
    order == "Apoidea",
    species_nm != "Apis mellifera",
    str_detect(sample_code, ".+_TS.+")
  ) |>
  group_by(location, month) |>
  summarise(n_ind = sum(no_ind, na.rm = TRUE))
```


