---
title: "Super simple power analysis for 2024 data"
author: "Emma Cartuyvels"
date: "2025-05-05"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---
# Intro

In this document I attempt to make a simple simulation and power analysis using the simr package.

We use the following model:
$$
n = m + y + (1|l)
$$
With $n$ being the number of individuals, $m$ the month, $y$ the year and $l$ the location.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(here)
library(tidyverse)
library(simr)
```

# Information from SPRING experiment

At this moment we only have a limited data set to determine our coefficients from.

```{r gegevens 2024}
identifications <- read_excel(here("data", "zandstreek_2024.xlsx"),
  sheet = "identifications"
)

identifications <- identifications |>
  janitor::clean_names() |>
  as_tibble() |>
  rename(no_ind = "mm_ff") |>
  mutate(
    location = substr(sample_code, 1, 6),
    month = substr(sample_code, 18, 19)
  )

n_ind_per_loc <- identifications |>
  filter(
    species_nm != "Apis mellifera",
    str_detect(sample_code, ".+_TS.+")
  ) |>
  group_by(location, month) |>
  summarise(n_ind = sum(no_ind, na.rm = TRUE)) |>
  # correct for shorter transects compared to proposed protocol
  mutate(n_ind = n_ind * 7)
```

```{r}
model1 <- glmer(n_ind ~ month + (1 | location),
  family = "poisson",
  data = n_ind_per_loc
)
summary(model1)

attr(VarCorr(model1)$location, "stddev")
```
```{r}
emmeans::emmeans(model1, ~month)
```

## Design Setup

```{r}
# Define design variables
generate_data <- function(n_locations) {
  # Design variables
  years <- as.numeric(1:6)
  months <- as.factor(4:9)
  locs <- paste0("loc_", 1:n_locations)

  # Base: 1 visit per month per year per location
  data <- expand.grid(
    location = locs,
    year = years,
    month = months,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  data$month <- factor(data$month, levels = levels(months))
  return(data)
}
```

## Model Specification

```{r}
# Fixed effects: Intercept, month 6-9, year (placeholder at index 7)
fixed_base <- c(0.991, 0,
                unname(coef(summary(model1))[1, 1]) - 0.991,
                unname(coef(summary(model1))[2:4, 1]),
                NA)

# Random effect variance for location
rand <- attr(VarCorr(model1)$location, "stddev")
varcorrmat1 <- list(location = c(rand^2))
```

## Power Analysis Function

```{r}
run_power_analysis <- function(year_slope_percent, n_locations, nsim = 100) {
  data <- generate_data(n_locations)

  # Convert percent trend to log scale
  year_slope <- log(1 + year_slope_percent / 100)
  fixed <- fixed_base
  fixed[7] <- year_slope

  # Simulated model with defined structure
  model <- makeGlmer(
    formula = n_ind ~ month + year + (1 | location),
    family = poisson,
    fixef = fixed,
    VarCorr = varcorrmat1,
    data = data
  )

  test <- powerSim(model,
    nsim = nsim, test = fixed("year", "z"),
    progress = FALSE
  )

  return(test)
}
```

## Run Power Simulations

```{r}
# Define trends and locations
trends <- c(1, 2)
locations <- c(seq(10, 100, 10), 125, 150, 175, 200)

# Run all combinations
results <- list()

for (t in trends) {
  for (l in locations) {
    message(sprintf("Running trend %.1f%% with %d locations", t, l))
    result <- run_power_analysis(t, l, nsim = 25)
    results[[paste0("trend", t, "_loc", l)]] <- result
  }
}
```

## Summarize Results

```{r}
# Print summaries
power_df <- bind_rows(
  lapply(names(results), function(name) {
    summary(results[[name]]) |>
      mutate(
        trend = str_extract(name, "([0-9]|\\.)+(?=_)"),
        points = as.numeric(str_extract(name, "[0-9]+$"))
      ) |>
      select(trend, points, everything())
  })
)

power_df
```

## Extract and Plot Power Results

```{r}
# Plot power curve
library(ggplot2)
ggplot(power_df, aes(x = points, y = mean, color = trend)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = trend, colour = NULL),
    alpha = 0.1
  ) +
  geom_point() +
  geom_hline(
    yintercept = 0.8,
    linetype = 2,
    color = "darkgrey"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Power to Detect Yearly Trends",
    x = "Number of Locations",
    y = "Power"
  ) +
  theme_minimal()
```

# Wild bees

```{r}
n_ind_per_loc <- identifications |>
  filter(
    order == "Apoidea",
    species_nm != "Apis mellifera",
    str_detect(sample_code, ".+_TS.+")
  ) |>
  group_by(location, month) |>
  summarise(n_ind = sum(no_ind, na.rm = TRUE))
```




