---
title: "Super simple power analysis for 2024 data"
author: "Emma Cartuyvels"
date: "2025-05-05"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---
# Intro

In this document I attempt to make a simple simulation and power analysis using the simr package.

We use the following model:
$$
n = m + y + (1|l)
$$
With $n$ being the number of individuals (modeled using a Poisson distribution), $m$ the month, $y$ the year and $l$ the location. Since we are using a random effect for location we will be using the `simr` package which allows us to create an artificial mixed model object.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(here)
library(tidyverse)
library(simr)
```

# Information from SPRING experiment

At this moment we only have a limited data set to determine our coefficients from.

```{r gegevens 2024}
bestuivers_2024 <- read_csv(here("data", "processed", "bestuivers_2024"))

bestuivers_2024 <- bestuivers_2024 |>
  rename(location = location_code,
         month = maand) |>
  filter(
    species_nm != "Apis mellifera",
    method_cd == "TS"
  ) |>
  group_by(location, month) |>
  summarise(n_ind = sum(n_ind, na.rm = TRUE)) |>
  # correct for shorter transects compared to proposed protocol
  # length new transects / length old transects = (1000 * 3) / (160 * 4) ~= 5
  mutate(n_ind = n_ind * 5,
         month = as.factor(month))
```

```{r}
model1 <- glmer(n_ind ~ month + (1 | location),
  family = "poisson",
  data = bestuivers_2024
)
summary(model1)

attr(VarCorr(model1)$location, "stddev")
```
```{r}
emmeans::emmeans(model1, ~month)
```

## Design Setup

```{r}
# Define design variables
generate_data <- function(n_locations, n_years) {
  # Design variables
  years <- as.numeric(1:n_years)
  months <- as.factor(4:9)
  locs <- paste0("loc_", 1:n_locations)

  # Base: 1 visit per month per year per location
  data <- expand.grid(
    location = locs,
    year = years,
    month = months,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  data$month <- factor(data$month, levels = levels(months))
  return(data)
}
```

## Model Specification

```{r}
#' Fixed effects: Intercept (month 4), month 5-9, year (placeholder at index 7)
#' Month 4 and 5 based on SPRING data
fixed_base <- c(1.674776, 0,
                unname(coef(summary(model1))[1, 1]) - 1.674776,
                unname(coef(summary(model1))[2:4, 1]) +
                  unname(coef(summary(model1))[1, 1]) - 1.674776,
                NA)

# Random effect variance for location and year
rand_loc <- attr(VarCorr(model1)$location, "stddev")
rand_year <- 0.15

varcorrmat1 <- list(
  location = c(rand_loc^2),
  year = c(rand_year^2)
)
```

## Power Analysis Function

```{r}
run_power_analysis <- function(year_slope_percent,
                               n_locations,
                               n_years,
                               nsim = 100) {
  data <- generate_data(n_locations, n_years)

  # Convert percent trend to log scale
  year_slope <- log(1 + year_slope_percent / 100)
  fixed <- fixed_base
  fixed[7] <- year_slope

  # Simulated model with defined structure
  model <- makeGlmer(
    formula = n_ind ~ month + year + (1 | location) + (1 | year),
    family = poisson,
    fixef = fixed,
    VarCorr = varcorrmat1,
    data = data
  )

  test <- powerSim(model,
    nsim = nsim, test = fixed("year", "z"),
    progress = FALSE
  )

  return(test)
}
```

## Run Power Simulations

```{r}
# Define trends and locations
trends <- c(0.5, 1, 2)
locations <- c(250, 500)
n_years_v  <- c(6, 8, 10, 12)

# Run all combinations
results <- list()

for (ny in n_years_v) {
  for (t in trends) {
    for (l in locations) {
      message(sprintf("n_years %d | trend %.1f%% | locations %d", ny, t, l))
      result <- run_power_analysis(t, l, n_years = ny, nsim = 500)
      results[[paste0("ny", ny, "_trend", t, "_loc", l)]] <- result
    }
  }
}
```

## Summarize Results

```{r}
# Print summaries
power_df <- bind_rows(
  lapply(names(results), function(name) {
    summary(results[[name]]) |>
      mutate(
        year  = as.numeric(str_extract(name, "(?<=ny)\\d+")),
        trend = str_extract(name, "(?<=trend)\\d+(?:\\.\\d+)?"),
        points = as.numeric(str_extract(name, "(?<=loc)\\d+"))
      ) |>
      select(year, trend, points, everything())
  })
)

power_df
```

## Extract and Plot Power Results

```{r}
# Plot power curve
library(ggplot2)
ggplot(power_df, aes(x = points, y = mean, color = trend)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = trend, colour = NULL),
    alpha = 0.1
  ) +
  geom_point() +
  geom_hline(
    yintercept = 0.8,
    linetype = 2,
    color = "darkgrey"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Power to Detect Yearly Trends",
    x = "Number of Locations",
    y = "Power"
  ) +
  theme_minimal() +
  facet_grid(~year)
```

# Wild bees

```{r, eval=FALSE}
n_ind_per_loc <- bestuivers_2024 |>
  filter(
    order == "Apoidea",
    species_nm != "Apis mellifera",
    str_detect(sample_code, ".+_TS.+")
  ) |>
  group_by(location, month) |>
  summarise(n_ind = sum(no_ind, na.rm = TRUE))
```


